Synthesizing Symmetric Lenses

### 2018-07-11 papers/SymmetricLenses/main.tex
add new run data and fill in concrete numbers

We describe an implementation of simple symmetric lenses and our synthesis
procedure as extensions to the Boomerang language~\cite{Boomerang}. We
evaluate
its performance on 45 benchmark examples drawn from FlashFill, Augeus,
and an online collection of electronic file formats. We find this implementation is
able to synthesize all the lenses of the benchmark suite in under 30 seconds.
\end{abstract}

@@

However, there is a price for narrowing the synthesis domain so
dramatically: Many real data formats are almost, but not quite,
bijections. More often two related data formats have some data in
common, but also contain some differences. For instance, two related
tables might share four of five columns, but each has a different
fifth column. Alternately, one ad hoc configuration file might include
some file-specific meta data, such as a date, a time or a reference
number, whereas the other does not. Indeed of the 45 benchmark problems
described in Section \ref{sec:evaluation}, all of the benchmarks
taken from Flash Fill~\cite{flashfill} and all of the benchmarks that
synchronize between two ad hoc file formats have this 
characteristic, \afm{num} benchmarks in total. Consequently, tools that search
only for strict bijections will rarely be applicable in practice.

@@

\begin{figure}
  \includegraphics[width=.5\textwidth]{high-level-algorithm.pdf}
  \caption{Schematic diagram for the simple symmetric lens synthesis
  algorithm. The user provides 
   regular expressions \BRegex and \BRegexAlt and a set of examples \Examples{}
    as input. \Expand first converts \BRegex and \BRegexAlt to
    stochastic regular expressions \Regex and \RegexAlt with default
    probabilities. It then finds pairs of stochastic regular expressions
    equivalent to \Regex and \RegexAlt and iteratively proposes them to \GreedySynth.
    \GreedySynth finds a lens typed between the supplied SREs.
    When the algorithm determines it has found a likely lens, it
  returns it.  \ksf{Can we change cost to likelihood and l\_likely to l\_best?}
   }
  \label{fig:high-level-algorithm}
\end{figure}

@@

We implemented the \GreedySynth and \Expand algorithms and evaluated their
effectiveness at synthesizing lenses on a 
benchmark suite of 45 problems, including data cleaning tasks taken
from Flash Fill~\cite{gulwani-popl-2014}, view maintenance
tasks taken from Augeas~\cite{Augeas}, and ad hoc format synchronization tasks
taken from a variety of other locations\footnote{The majority of these format
  synchronization tasks were found from the following index:
  \url{http://fileformats.archiveteam.org/wiki/Electronic_File_Formats}}.

In summary, our main contributions are:
\begin{itemize}
\item We define simple symmetric lenses, which have
  the useful property that they do not maintain any hidden state
  (\S\ref{sec:overview}).  
  We give a language for defining such lenses and describe how to
  calculate their likelihoods (\S\ref{sec:ssl}). 
\item  We show that simple symmetric lenses
  fall between asymmetric lenses~\cite{Focal2005-long2} and full
  symmetric lenses~\cite{symmetric-lenses} in expressiveness
  (\S\ref{sec:related}).  We further show that the class of simple
  lenses can be characterized semantically as the subset of full
  symmetric lenses for which a simple ``forgetfulness'' property holds
  (\S\ref{sec:relationship}).
\item We show how to synthesize simple symmetric lenses (including all
  asymmetric lenses as a special case), using a novel application of stochastic
  regular expressions to guide the search process (\S\ref{sec:synthesis}). 
  To support that effort, we show that the star semi-ring equivalences on regular
  expressions remain valid when applied to stochastic regular
  expressions, preserving the probability distributions as well as the languages
  (\S\ref{sec:sre}).
\item We extend Boomerang with simple symmetric lenses,
  showing that such lenses can integrate smoothly with 
  other lens extensions such as quotient lenses and matching
  lenses.  We evaluate our
  implementation on a set of 45 lens synthesis benchmarks drawn from
  data cleaning, view maintenance, 
  and file synchronization tasks.  We demonstrate that we can
  synthesize simple symmetric lenses for all of the benchmarks in
  under 30 seconds (\S\ref{sec:evaluation}). 
\end{itemize}
We finish the paper with a discussion of related work (\S\ref{sec:related}) and
our concluding thoughts (\S\ref{sec:conc}).

@@

    $\REntropyOf{\Regex \Given \InvertOf{\Lens}, \RegexAlt}$
    & =
    & $\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$\\
  \end{tabular}
\end{center}
$\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$ is defined symmetrically. These
functions bound the expected number of bits to recover one data format from a
synchronized string in the other format. Note that we would be able to exactly
calculate the conditional entropy, were it not for \MergeL. If
$\MergeLOf{\Lens_1}{\Lens_2} \OfType \Regex \Leftrightarrow
\PRegexOr{\RegexAlt_1}{\RegexAlt_2}{\ProbabilityAlt}$, given a string in $s$, we
need to determine if the synchronized string is in $\RegexAlt_1$ or $\RegexAlt_2$.
However, this information content is dependent on the how likely the
synchronized string is to be in $\RegexAlt_1$ or $\RegexAlt_2$. Nevertheless, we
typically calculate the conditional entropy exactly, as merges are relatively
uncommon on practice; only \afm{2} of the lenses sythesized in our benchmarks
include merges.

@@

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
Algorithm~\ref{alg:synth-sym-lens} presents our synthesis algorithm at a high
level of abstraction. The input regular expressions are first converted into
stochastic regular expressions with \ToStochastic.  
%Once that conversion has
%taken place, the $\RXSearch$ algorithm executes. This algorithm itself is
%defined via five functions: (1) $\PCF{\PQ.Create}$, (2) $\PCF{\PQ.Pop}$, (3)
%$\PCF{Expand}$, (4) $\PCF{\PQ.Push}$, and (5) $\PCF{\Continue}$. First,
This pair of SREs is used to initialize a priority queue ($pq$).
Next, $\PCF{SynthSymLens}$
enters a loop that searches for likely lenses. The loop terminates when the
algorithm believes it is unlikely to find a better lens than the best one it has
found so far (a termination condition defined by $\PCF{\Continue}$). Within each
iteration of the loop, it:
\begin{itemize}
\item pops the next class ($S$, $T$) of lenses to
search off of the priority queue ($\PCF{\PQ.Pop}$),
\item executes $\GreedySynth$ to find a best lens in that class if
one exists ($\LC$),
using the examples $exs$ to filter out potential lenses that do not satisfy
the specification,
\item replaces $best$ with $\LC$, if $\LC$ has lower cost according
to our information theoretic metric, and
\item adds the SREs derived from rewriting $S$ and $T$ ($\Expand(S,T)$)
  to the priority queue.  The priority of a SRE pair is the
  number of rewrites needed to derive the pair from the originals.
\end{itemize}
When the loop terminates, the search returns the globally best lens found
($best$).  Each subroutine of this algorithm will be explained in
further depth in the following subsections.

\begin{algorithm}[t]
  \caption{\SynthSymLens}
  \label{alg:synth-sym-lens}
  \begin{algorithmic}[1]
    \Function{SynthSymLens}{$\BRegex,\BRegexAlt,\Examples$}
    \State $\Regex \gets \Call{\ToStochastic}{\BRegex}$
    \State $\RegexAlt \gets \Call{\ToStochastic}{\BRegexAlt}$
    \State $\RXSearchState \gets \Call{\PQ.Create}{\Regex,\RegexAlt}$
    \State $\Best \gets \None$
    \While{$\Call{\Continue}{\RXSearchState,\Best}$}
    \State $(\Regex,\RegexAlt) \gets \Call{\PQ.Pop}{\RXSearchState}$
    \State $\LC \gets \Call{\GreedySynth}{\Examples,\Regex,\RegexAlt}$
    \If{$\Call{Cost}{\LC} < \Call{Cost}{\Best}$}
    \State $\Best \gets \LC$
    \EndIf
    \State $\Call{\PQ.Push}{\RXSearchState,\Expand(\Regex,\RegexAlt)}$
    \EndWhile
    \State \ReturnVal{best}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Search Through Types}
The first synthesizer is one that searches through regular expression pairs.
This synthesizer comprises \Expand, \PQ, and \Continue. It proposes regular
expressions that are semantically, but not syntactically, equivalent to the
originals by performing the following rewrites:
\begin{center}
  \begin{tabular}{rcl}
    $\PRegexStar{\Regex}{\Probability}$
    & \Rewrite
    & \PRegexOr{\EmptyString{}}{(\RegexConcat{\Regex{}}{\PRegexStar{\Regex{}}{\Probability}})}{\Probability}\\

@@

The priority queue yields stochastic regular expressions generated using fewer
rewrites first. Only when there are no more proposed regular expressions derived
from $n$ rewrites will \PCF{\PQ.Pop} propose regular expressions derived from
$n+1$ rewrites.

@@

When $\Call{Continue}{pq}$ returns $\False$, \Expand stops proposing regular
expression pairs, and instead returns to the user the best lens found thus far.
Note this means that if the algorithm finds a bijective lens, which has zero cost, it
will immediately return.
%
\subsection{Stochastic DNF Regular Expressions}
Stochastic DNF regular expressions (\SDNFREabbrev) are a pseudo-normal form
for stochastic regular expressions.  Of all the star-semiring equations
required to relate two \SDNFREabbrevs, only those relating to Kleene star
(\EG, \emph{Unrollstar$_L$} and \emph{Unrollstar$_R$}) are needed to relate
two \SDNFREabbrevs; other axioms, like associativity, commutativity
or distributivity, are unneeded.  This important property makes it possible to
construct an efficient algorithm ($\GreedySynth$), directed
by the syntax of two \SDNFREabbrevs, to synthesize any lenses in the
same ``class'' (\IE, any lenses that have the same type when the
Kleene star axioms are unused).  This property, and the resulting
efficiency of $\GreedySynth$ is key to the effectiveness of our
core algorithms.

@@

Intuitively, stochastic DNF regular expressions are stochastic
regular expressions with all concatenations fully distributed over all
disjunctions. As such, the language of a stochastic DNF regular expression is a
union of its subcomponents, the language of a stochastic sequence is the
concatenation of its subcomponents, and the language of a stochastic atom is the
iteration of its subcomponent. For
$\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)}$
to be a valid stochastic DNF regular expression, the probabilities must sum to
one ($\sum_{i=0}^n\Probability_i = 1$).

@@

As these DNF regular expressions are \emph{stochastic}, they are annotated with
probabilities to express a probability distribution, in addition to a
language. \saz{I find the use of asterisk for the $\times$ or $\cdot$ or maybe
  just juxtaposition slightly jarring.}
\begin{center}
  \begin{tabular}{rcl}
    $\ProbabilityOf{\PRegexStar{\DNFRegex}{\Probability}}{\String}$
    & =
    & $\Sigma_n \Sigma_{\String = \String_1 \ldots \String_n}\Probability^n*(1-\Probability)*\Pi_{i=1}^n\ProbabilityOf{\DNFRegex}{\String_i}$\\

@@

The algorithm for converting a stochastic regular expressions into its
corresponding stochastic DNF regular expressions, written
$\ToDNFRegexOf{\Regex}$, is defined below.
\[
  \begin{array}{rcl}
    \ToDNFRegexOf{\String} & = & \DNFOf{(\SequenceOf{\String},1)}\\
    \ToDNFRegexOf{\emptyset} & = & \DNFOf{}\\
    \ToDNFRegexOf{(\PRegexStar{\Regex}{\Probability})} & = & \AtomToDNFOf{\PRegexStar{(\ToDNFRegexOf{\Regex})}{\Probability}}\\
    \ToDNFRegexOf{(\RegexConcat{\Regex_1}{\Regex_2})} & = & \ToDNFRegexOf{\Regex_1} \ConcatDNF \ToDNFRegexOf{\Regex_2}\\
    \ToDNFRegexOf{(\PRegexOr{\Regex_1}{\Regex_2}{\Probability})} & = & \ToDNFRegexOf{\Regex_1} \OrDNF_{\Probability} \ToDNFRegexOf{\Regex_2}\\
  \end{array}
\]
This conversion relies on operators defined in
Figure~\ref{fig:dnf-regex-functions}. We have proved that this conversion
respects languages and probability distributions.

@@

\paragraph*{\ToStochastic} With stochastic DNF regular expressions and
\ToDNFRegex defined, it is easier to explain \ToStochastic, the function that converts regular
expressions into stochastic regular expressions. If users
provided stochastic regular expressions themselves (perhaps inferred from
example data), this step could be skipped,
though our implementation does not (yet) permit users to provide such
stochastic regular expressions. If $\Regex$ is a stochastic DNF regular
expression generated by \ToStochastic, then $\ToDNFRegexOf{\Regex} =
\DNFOf{(\Sequence_1,\frac{1}{n}) \DNFSep \ldots \DNFSep
  (\Sequence_n,\frac{1}{n})}$ for some sequences $\Sequence_1\ldots\Sequence_n$,
and every stochastic atom generated by \ToStochastic is
$\PRegexStar{\DNFRegex}{.8}$. In particular, $\ToDNFRegex$ generates regular
expressions whose DNF form gives equal probability to all sequence subcomponents
of the DNF regular expression., and gives iterations a .8 chance to continue
iterating. In our experience generating random strings from regular expressions,
these probabilities provide good distributions of strings---stars are iterated
4 times on average, and no individual choice in a series disjunctions is chosen
disproportionately often. As we have found these distributions generate
reasonable strings, we believe they describe the shape of data well.

@@

\subsection{\GreedySynth}
\label{subsec:greedy-synth}
The synthesis procedure comprises three algorithms: one that greedily finds
symmetric DNF lenses (\GreedySynth), one that greedily finds symmetric sequence
lenses (\GreedySeqSynth), and one that finds symmetric atom lenses
(\AtomSynth{}). These three algorithms are hierarchically structured:
\GreedySynth relies on \GreedySeqSynth, \GreedySeqSynth relies on \AtomSynth,
and \AtomSynth rlies on \GreedySynth. The structure of the algorithms mirrors
the structure of symmetric DNF lenses and \SDNFREabbrevs. We describe this
synthesis algorithm using the running example of searching for a lens between
\lstinline{"" | name.name*} and \lstinline{"" | name}.

\paragraph*{Symmetric DNF Lenses} Algorithm~\ref{alg:greedysynth} presents
\GreedySynth, which synthesizes symmetric DNF lenses. Its inputs 
are a suite of input-output examples and a pair of stochastic DNF regular expressions.
%
First, \PCF{CannotMap} determines whether there is no lens satisfying the
examples, and if so, \GreedySynth returns \None immediately. Otherwise, \GreedySynth finds
the best lenses (given the examples that match them) between all sequence pairs
$(SQ_i, TQ_j)$ drawn from the left and right DNF regular expressions. (The
function \CartesianMap maps its argument across the cross product of the input
lists). A priority queue containing these sequence lenses, ordered by cost, is
then initialized with $\PCF{PQ.Create}$. The symmetric lens is then built up
iteratively from these sequence lenses, where the state of the partially
constructed lens is tracked in the lens builder, $lb$.

\GreedySynth loops until there are no more sequence lenses in the priority
queue. Within this loop, a sequence lense is popped from the queue and, if it is
``useful,'' it is included in the final DNF lens. The lens is considered to
be useful when its source (or target) is \textit{not} already the source (or target) of
an already chosen sequence lens. If the examples require that two sequences have a
lens between them, then any sequence lens between them is considered useful. We
update the priorities of the sequence lenses as the algorithm proceeds: if two
sequence lenses have the same source, the second one to be popped gets a higher
cost than it originally had, to account for the information that needs to be stored
for including that source of non-bijectivity.

@@

\begin{algorithm}[t]
  \caption{\GreedySynth}
  \label{alg:greedysynth}
  \begin{algorithmic}[1]
    \Function{GreedySynth}{$\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}$}
    \If{$\Call{CannotMap}{\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}}$}
    \State $\ReturnVal{\None}$
    \EndIf
    \State $\SLS \gets
    \Call{\CartesianMap}{\GreedySeqSynth(\Examples),\ListOf{\Sequence_1;\ldots;\Sequence_n},\ListOf{\SequenceAlt_1;\ldots;\SequenceAlt_m}}$
    \State $\GreedyState \gets \Call{PQ.Create}{\SLS}$
    \State $lb \gets \Call{LensBuilder.Empty}{}$
    \While{$\Call{PQ.IsNonempty}{\GreedyState}$}
    \State $\SSQLens \gets
    \Call{PQ.Pop}{\GreedyState}$
    \If{$\Call{LensBuilder.UsefulAdd}{lb,\SSQLens,\Examples}$}
    \State $lb \gets \Call{LensBuilder.AddSeq}{lb,\SSQLens}$
    \EndIf
    \EndWhile
    \State \ReturnVal{\Call{LensBuilder.ToDNFLens}{\GreedyState}}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\paragraph*{Symmetric Sequence lenses} Algorithm~\ref{alg:greedyseqsynth}
presents \GreedySeqSynth, which synthesizes symmetric sequence lenses using an
algorithm whose structure is similar to \GreedySynth{}'s.  It calls \PCF{AtomSynth}, 
which synthesizes atom lenses by iterating a DNF lens between
its subcomponents.

>>>>

We describe an implementation of simple symmetric lenses and our synthesis
procedure as extensions to the Boomerang language~\cite{Boomerang}. We
evaluate
its performance on 48 benchmark examples drawn from FlashFill, Augeus,
and an online collection of electronic file formats. We find this implementation is
able to synthesize all the lenses of the benchmark suite in under 30 seconds.
\end{abstract}

@@

However, there is a price for narrowing the synthesis domain so
dramatically: Many real data formats are almost, but not quite,
bijections. More often two related data formats have some data in
common, but also contain some differences. For instance, two related
tables might share four of five columns, but each has a different
fifth column. Alternately, one ad hoc configuration file might include
some file-specific meta data, such as a date, a time or a reference
number, whereas the other does not. Indeed of the 48 benchmark problems
described in Section \ref{sec:evaluation}, all of the benchmarks
taken from Flash Fill~\cite{flashfill} and all of the benchmarks that
synchronize between two ad hoc file formats have this 
characteristic, 48 benchmarks in total. Consequently, tools that search
only for strict bijections will rarely be applicable in practice.

@@

\begin{figure}
  \includegraphics[width=.5\textwidth]{high-level-algorithm.pdf}
  \caption{Schematic diagram for the simple symmetric lens synthesis algorithm.
    The user provides regular expressions \BRegex and \BRegexAlt and a set of
    examples \Examples{} as input. \Expand first converts \BRegex and \BRegexAlt
    to stochastic regular expressions \Regex and \RegexAlt with default
    probabilities. It then finds pairs of stochastic regular expressions
    equivalent to \Regex and \RegexAlt and iteratively proposes them to
    \GreedySynth. \GreedySynth finds a lens typed between the supplied SREs.
    When the algorithm determines it has found a likely lens, it returns it.}
  \label{fig:high-level-algorithm}
\end{figure}

@@

We implemented the \GreedySynth and \Expand algorithms and evaluated their
effectiveness at synthesizing lenses on a benchmark suite of 48 problems,
including data cleaning tasks taken from Flash Fill~\cite{gulwani-popl-2014},
view maintenance tasks taken from Augeas~\cite{Augeas}, and ad hoc format
synchronization tasks taken from a variety of other locations\footnote{The
  majority of these format synchronization tasks were derived from the
  bidirectional programming literature.}.

In summary, our main contributions are:
\begin{itemize}
\item We define simple symmetric lenses, which have
  the useful property that they do not maintain any hidden state
  (\S\ref{sec:overview}).  
  We give a language for defining such lenses and describe how to
  calculate their likelihoods (\S\ref{sec:ssl}). 
\item  We show that simple symmetric lenses
  fall between asymmetric lenses~\cite{Focal2005-long2} and full
  symmetric lenses~\cite{symmetric-lenses} in expressiveness
  (\S\ref{sec:related}).  We further show that the class of simple
  lenses can be characterized semantically as the subset of full
  symmetric lenses for which a simple ``forgetfulness'' property holds
  (\S\ref{sec:relationship}).
\item We show how to synthesize simple symmetric lenses (including all
  asymmetric lenses as a special case), using a novel application of stochastic
  regular expressions to guide the search process (\S\ref{sec:synthesis}). 
  To support that effort, we show that the star semi-ring equivalences on regular
  expressions remain valid when applied to stochastic regular
  expressions, preserving the probability distributions as well as the languages
  (\S\ref{sec:sre}).
\item We extend Boomerang with simple symmetric lenses,
  showing that such lenses can integrate smoothly with 
  other lens extensions such as quotient lenses and matching
  lenses.  We evaluate our
  implementation on a set of 48 lens synthesis benchmarks drawn from
  data cleaning, view maintenance, 
  and file synchronization tasks.  We demonstrate that we can
  synthesize simple symmetric lenses for all of the benchmarks in
  under 30 seconds (\S\ref{sec:evaluation}). 
\end{itemize}
We finish the paper with a discussion of related work (\S\ref{sec:related}) and
our concluding thoughts (\S\ref{sec:conc}).

@@

    $\REntropyOf{\Regex \Given \InvertOf{\Lens}, \RegexAlt}$
    & =
    & $\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$\\
  \end{tabular}
\end{center}
$\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$ is defined symmetrically. These
functions bound the expected number of bits to recover one data format from a
synchronized string in the other format. Note that we would be able to exactly
calculate the conditional entropy, were it not for \MergeL. If
$\MergeLOf{\Lens_1}{\Lens_2} \OfType \Regex \Leftrightarrow
\PRegexOr{\RegexAlt_1}{\RegexAlt_2}{\ProbabilityAlt}$, given a string in $s$, we
need to determine if the synchronized string is in $\RegexAlt_1$ or $\RegexAlt_2$.
However, this information content is dependent on the how likely the
synchronized string is to be in $\RegexAlt_1$ or $\RegexAlt_2$. Nevertheless, we
typically calculate the conditional entropy exactly, as merges are relatively
uncommon on practice; only 2 of the lenses sythesized in our benchmarks
include merges.

@@

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
Algorithm~\ref{alg:synth-sym-lens} presents our synthesis algorithm at a high
level of abstraction. The input regular expressions are first converted into
stochastic regular expressions with \ToStochastic.  
%Once that conversion has
%taken place, the $\RXSearch$ algorithm executes. This algorithm itself is
%defined via five functions: (1) $\PCF{\PQ.Create}$, (2) $\PCF{\PQ.Pop}$, (3)
%$\PCF{Expand}$, (4) $\PCF{\PQ.Push}$, and (5) $\PCF{\Continue}$. First,
This pair of SREs is used to initialize a priority queue ($pq$).
Next, $\PCF{SynthSymLens}$
enters a loop that searches for likely lenses. The loop terminates when the
algorithm believes it is unlikely to find a better lens than the best one it has
found so far (a termination condition defined by $\PCF{\Continue}$). Within each
iteration of the loop, it:
\begin{itemize}
\item pops the next class ($S$, $T$) of lenses to
search off of the priority queue ($\PCF{\PQ.Pop}$),
\item executes $\GreedySynth$ to find a best lens in that class if
one exists ($\LC$),
using the examples $exs$ to filter out potential lenses that do not satisfy
the specification,
\item replaces $best$ with $\LC$, if $\LC$ has lower cost according
to our information theoretic metric, and
\item adds all of the SREs derived from rewriting $S$ and $T$ ($\Expand(S,T)$)
  to the priority queue with $\PCF{\PQ.Push}$. The priority of a SRE pair is the
  number of rewrites needed to derive the pair from the originals.
\end{itemize}
When the loop terminates, the search returns the globally best lens found
($best$).  Each element of this algorithm will be explained in
further depth in the following subsections.

\begin{algorithm}
  \caption{\SynthSymLens}
  \label{alg:synth-sym-lens}
  \begin{algorithmic}[1]
    \Function{SynthSymLens}{$\BRegex,\BRegexAlt,\Examples$}
    \State $\Regex \gets \Call{\ToStochastic}{\BRegex}$
    \State $\RegexAlt \gets \Call{\ToStochastic}{\BRegexAlt}$
    \State $\RXSearchState \gets \Call{\PQ.Create}{\Regex,\RegexAlt}$
    \State $\Best \gets \None$
    \While{$\Call{\Continue}{\RXSearchState,\Best}$}
    \State $(\Regex,\RegexAlt) \gets \Call{\PQ.Pop}{\RXSearchState}$
    \State $\LC \gets \Call{\GreedySynth}{\Examples,\Regex,\RegexAlt}$
    \If{$\Call{Cost}{\LC} < \Call{Cost}{\Best}$}
    \State $\Best \gets \LC$
    \EndIf
    \State $\Call{\PQ.Push}{\RXSearchState,\Expand(\Regex,\RegexAlt)}$
    \EndWhile
    \State \ReturnVal{best}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Search Through Types}
The first synthesizer is one that searches through regular expression pairs.
This synthesizer comprises \Expand, \PQ, and \Continue. It proposes regular
expressions that are semantically, but not syntactically, equivalent to the
originals. In particular, this synthesizer performs the following rewrites:
\begin{center}
  \begin{tabular}{rcl}
    $\PRegexStar{\Regex}{\Probability}$
    & \Rewrite
    & \PRegexOr{\EmptyString{}}{(\RegexConcat{\Regex{}}{\PRegexStar{\Regex{}}{\Probability}})}{\Probability}\\

@@

Our priority queue proposes stochastic regular expressions generated from fewer
rewrites first. Only when there are no more proposed regular expressions derived
from $n$ rewrites will \PCF{\PQ.Pop} propose regular expressions derived from
$n+1$ rewrites.

@@

When $\Call{Continue}{pq}$ returns $\False$, \RXSearch stops proposing regular
expression pairs, and instead returns to the user the best lens found thus far.
Note this means that if \RXSearch finds a bijective lens, which has zero cost it
will immediately return.
%
\subsection{Stochastic DNF Regular Expressions}
Stochastic DNF regular expressions (\SDNFREabbrev) are a pseudo-normal form
for stochastic regular expressions.  Of all the star-semiring equations
required to relate two \SDNFREabbrevs, only those relating to Kleene star
(\EG, \emph{Unrollstar$_L$} and \emph{Unrollstar$_R$}) are needed to relate
two \SDNFREabbrevs, other axioms, like associativity, commutativity
or distributivity, are unneeded.  This important property makes it possible to
construct an efficient algorithm ($\GreedySynth$), directed
by the syntax of two \SDNFREabbrevs, to synthesize any lenses in the
same ``class'' (\IE, any lenses that have the same type when the
Kleene star axioms are unused).  This property, and the resulting
efficiency of $\GreedySynth$ is key to the effectiveness of our
core algorithms.

@@

Intuitively, stochastic DNF regular expressions are essentially stochastic
regular expressions with all concatenations fully distributed over all
disjunctions. As such, the language of a stochastic DNF regular expression is a
union of its subcomponents, the language of a stochastic sequence is the
concatenation of its subcomponents, and the language of a stochastic atom is the
iteration of its subcomponent. For
$\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)}$
to be a valid stochastic DNF regular expression, the probabilities must sum to
one ($\sum_{i=0}^n\Probability_i = 1$).

@@

As these DNF regular expressions are \emph{stochastic}, they are annotated with
probabilities to express a probability distribution, in addition to a language. 
\begin{center}
  \begin{tabular}{rcl}
    $\ProbabilityOf{\PRegexStar{\DNFRegex}{\Probability}}{\String}$
    & =
    & $\Sigma_n \Sigma_{\String = \String_1 \ldots \String_n}\Probability^n*(1-\Probability)*\Pi_{i=1}^n\ProbabilityOf{\DNFRegex}{\String_i}$\\

@@

We have a means of turning stochastic regular expressions into stochastic DNF
regular expressions. The conversion algorithm itself, written
$\ToDNFRegexOf{\Regex}$, is defined below.
\[
  \begin{array}{rcl}
    \ToDNFRegexOf{\String} & = & \DNFOf{(\SequenceOf{\String},1)}\\
    \ToDNFRegexOf{\emptyset} & = & \DNFOf{}\\
    \ToDNFRegexOf{(\PRegexStar{\Regex}{\Probability})} & = & \AtomToDNFOf{\PRegexStar{(\ToDNFRegexOf{\Regex})}{\Probability}}\\
    \ToDNFRegexOf{(\RegexConcat{\Regex_1}{\Regex_2})} & = & \ToDNFRegexOf{\Regex_1} \ConcatDNF \ToDNFRegexOf{\Regex_2}\\
    \ToDNFRegexOf{(\PRegexOr{\Regex_1}{\Regex_2}{\Probability})} & = & \ToDNFRegexOf{\Regex_1} \OrDNF_{\Probability} \ToDNFRegexOf{\Regex_2}\\
  \end{array}
\]
This conversion utilizes operators defined in
Figure~\ref{fig:dnf-regex-functions}. We have proved that this conversion
respects languages and probability distributions.

@@

\paragraph*{\ToStochastic} With stochastic DNF regular expressions and
\ToDNFRegex defined, it is easier to explain the function that converts regular
expressions into stochastic regular expressions, \ToStochastic. If users
specified stochastic regular expressions themselves (perhaps inferred from
example data), this step could be skipped,
though our implementation does not (yet) permit users to provide such
stochastic regular expressions. If $\Regex$ is a stochastic DNF regular
expression generated by \ToStochastic, then $\ToDNFRegexOf{\Regex} =
\DNFOf{(\Sequence_1,\frac{1}{n}) \DNFSep \ldots \DNFSep
  (\Sequence_n,\frac{1}{n})}$ for some sequences $\Sequence_1\ldots\Sequence_n$,
and every stochastic atom generated by \ToStochastic is
$\PRegexStar{\DNFRegex}{.8}$. In particular, $\ToDNFRegex$ generates regular
expressions whose DNF form gives equal probability to all sequence subcomponents
of the DNF regular expression., and gives iterations a .8 chance to continue
iterating. In our experience generating random strings from regular expressions,
these probabilities provide good distributions of strings---stars are iterated
4 times on average, and no individual choice in a series disjunctions is chosen
disproportionately often. As we have found these distributions generate
reasonable strings, we believe they describe the shape of data well.

@@

\subsection{\GreedySynth}
\label{subsec:greedy-synth}
The greedy algorithm is actually three algorithms, one that greedily finds
symmetric DNF lenses (\GreedySynth), one that greedily finds symmetric sequence
lenses (\GreedySeqSynth), and one that finds symmetric atom lenses
(\AtomSynth{}). These three algorithms are hierarchically structured:
\GreedySynth relies on \GreedySeqSynth, \GreedySeqSynth relies on \AtomSynth,
and \AtomSynth rlies on \GreedySynth. The structure of the algorithms mirrors
the structure of symmetric DNF lenses and \SDNFREabbrevs. We describe this
synthesis algorithm in the context of trying to find a lens between
\lstinline{"" | name.name*} and \lstinline{"" | name}.

\paragraph*{Symmetric DNF Lenses} Algorithm~\ref{alg:greedysynth} presents
\GreedySynth, a means to synthesize symmetric DNF lenses. The inputs to
\GreedySynth are a suite of input-output examples and a pair of stochastic DNF regular expressions.

First, \PCF{CannotMap} checks whether there is no possible lens satisfying the
examples. If so, \GreedySynth returns \None early. Otherwise, \GreedySynth finds
the best lenses (given the examples that match them) between all sequence pairs
$(SQ_i, TQ_j)$ drawn from the left and right DNF regular expressions. (The
function \CartesianMap maps its argument across the cross product of the input
lists). A priority queue containing these sequence lenses, ordered by cost, is
then initialized with $\PCF{PQ.Create}$. The symmetric lens is then built up
iteratively from these sequence lenses, where the state of the partially
constructed lens is tracked in the lens builder, $lb$.

\GreedySynth loops until there are no more sequence lenses in the priority
queue. Within this loop, a sequence lense is popped from the queue and, if it is
``useful'', then it is included in the final DNF lens. The lens is considered to
be useful when its source (or target) is not already the source (or target) of
an already chosen sequence lens. If the examples require that two sequences have a
lens between them, then any sequence lens between them is considered useful. We
update the priorities of the sequence lenses as the algorithm proceeds: if two
sequence lenses have the same source, the second one to be popped gets a higher
cost than it originally had, to account for the information that needs to be stored
for including that source of non-bijectivity.

@@

\begin{algorithm}
  \caption{\GreedySynth}
  \label{alg:greedysynth}
  \begin{algorithmic}[1]
    \Function{GreedySynth}{$\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}$}
    \If{$\Call{CannotMap}{\Examples,\DNFOf{(\Sequence_1,\Probability_1)\DNFSep\ldots\DNFSep(\Sequence_n,\Probability_n)},\DNFOf{(\SequenceAlt_1,\ProbabilityAlt_1)\DNFSep\ldots\DNFSep(\SequenceAlt_m,\ProbabilityAlt_m)}}$}
    \State $\ReturnVal{\None}$
    \EndIf
    \State $\SLS \gets
    \Call{\CartesianMap}{\GreedySeqSynth(\Examples),\ListOf{\Sequence_1;\ldots;\Sequence_n},\ListOf{\SequenceAlt_1;\ldots;\SequenceAlt_m}}$
    \State $\GreedyState \gets \Call{PQ.Create}{\SLS}$
    \State $lb \gets \Call{LensBuilder.Empty}{}$
    \While{$\Call{PQ.IsNonempty}{\GreedyState}$}
    \State $\SSQLens \gets
    \Call{PQ.Pop}{\GreedyState}$
    \If{$\Call{LensBuilder.UsefulAdd}{lb,\SSQLens,\Examples}$}
    \State $lb \gets \Call{LensBuilder.AddSeq}{lb,\SSQLens}$
    \EndIf
    \EndWhile
    \State \ReturnVal{\Call{LensBuilder.ToDNFLens}{\GreedyState}}
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\paragraph*{Symmetric Sequence lenses} Algorithm~\ref{alg:greedyseqsynth}
presents \GreedySeqSynth, which synthesizes symmetric sequence lenses using an
algorithm whose structure is similar to \GreedySynth{}'s.  It calls \PCF{AtomSynth}, a
function that synthesizes atom lenses by iterating a DNF lens between
its subcomponents.
